### 2021-12-07 text/outline/draft.tex
finish draft of data analysis section

Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}). Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated. Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary for both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie. We then applied the thresholding method to analyze both datasets, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.

% We compared GLM-EIV to thresholded regression on the real data and surprisingly found that the two methods produced similar results. Figures \ref{fig:real_data}a and \ref{fig:real_data}b show the estimate of the target of inference $\exp(\beta^m_1)$

To better understand this result, we examined the estimated effect of the perturbation on gRNA expression, $\exp(\hat{\beta}_1^g)$; this parameter plays an important role in determining problem difficulty. The value of $\exp(\hat{\beta}_1^g)$ was greater than expected: the 95\% CI for this parameter averaged across pairs was $[XX,XX]$ and $[300, 316]$ on the Gasperini and Xie data, respectively. We concluded that both datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).

To illuminate the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets. To this end, we sampled gRNA counts from a realistic model, holding fixed the raw gene expressions, library sizes, covariates, and fitted perturbation probabilities. We defined a normalized parameter called ``excess background contamination'' that controlled the problem difficulty. When excess background contamination was equal to zero, we sampled gRNA counts directly from the fitted model. As excess background contamination increased, we sampled gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells was increased while the mean gRNA expression of \textit{perturbed} cells was held constant. Finally, when excess background contamination was equal to $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sampled from the resulting model.

We generated partially-synthetic data for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We applied GLM-EIV and the thresholding method to analyze these data. We present results on two representative pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d. First, we observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (trend present across all genes; average relative difference $0.003$). This indicates that the fitted model accurately tracked the gRNA count distribution. Next, we observed that, as excess background contamination increased, the thresholding method exhibited severe attenuation bias, whereas GLM-EIV remained roughly stable.

We conducted a similar analysis on the entire set of positive control pairs. First, for each pair, we computed the ``relative estimate change'' (REC), defined as the relative difference between the estimate at some level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). % We plotted median

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Fold change estimates produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) aggregated across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV was more stable than thresholded regression as background contamination increased.}
	\label{fig:real_data}
\end{figure}

>>>>

Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}). Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated. Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary on both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie. We then applied the thresholding method to the data, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.

We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association). We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant. We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2\%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted). The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[XX,XX]$ and $[300,316]$ on Gasperini and Xie data, respectively). We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).

To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets. To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated. When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model. As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant. Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model. We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.

We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data. We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d. We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (result replicated across nearly all pairs; average relative difference $0.003$). This suggested that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets. Next, we observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable on the genes \textit{LRIF1} and \textit{NDUFA2}.

We generalized the above analysis to the entire set of positive control pairs. First, for each pair, we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC). As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable. Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate). GLM-EIV exhibited significantly higher CI coverage rates than thresholded regression as the data became increasingly contaminated.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}
	\label{fig:real_data}
\end{figure}
